---
title: "Harm and Economic Consequences of Severe Weather Events"
author: "Julio Costa"
date: "3/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

This analysis  explore the NOAA Storm Database to determine which types of severe weather  events are most harmful with respect to population health and which type of events have the greatest economic consequences.


# Data Processing
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. 

## Reading the data
We first read the data ensuring to use cache option to speed up the process.  We use the option not convert strings as factor as this makes a bit easier in analyzing some of the fields.

```{r,cache=TRUE,tidy=FALSE}
data<-read.csv("repdata_data_StormData.csv.bz2",stringsAsFactors = F)
```

We ensure the file size has the correct dimensions as expected and take a look at first rows.
```{r}

dim(data)
head(data)
names(data)

```



## Tidying up the data 

After analyzing the data we remove outliers and format it properly before running any results.

We create a subset of data to  keep only records with any property or damage cost and any  fatalities and injuries .
```{r,cache=TRUE,tidy=FALSE}
library(dplyr)
df<-select(data,BGN_DATE,EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
df.f<-filter(df,CROPDMG>0 | PROPDMG>0 | FATALITIES>0 | INJURIES>0)

```

we convert the exponential values to decimal.  There may be a more elegant weay of doing this.

```{r,tidy=FALSE}
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="K"]<-as.numeric(10^3)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="M"]<-as.numeric(10^6)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="B"]<-as.numeric(10^9)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="m"]<-as.numeric(10^6)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)==""]<-as.numeric(0)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="+"]<-as.numeric(1)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="0"]<-as.numeric(10)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="5"]<-as.numeric(10)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="6"]<-as.numeric(10)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="4"]<-as.numeric(10)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="h"]<-as.numeric(100)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="2"]<-as.numeric(10)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="7"]<-as.numeric(10)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="3"]<-as.numeric(10)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="H"]<-as.numeric(100)
df.f$PROPDMGEXP[(df.f$PROPDMGEXP)=="-"]<-as.numeric(0)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)==""]<-as.numeric(0)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)=="M"]<-as.numeric(10^6)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)=="K"]<-as.numeric(10^3)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)=="m"]<-as.numeric(10^6)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)=="B"]<-as.numeric(10^9)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)=="?"]<-as.numeric(0)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)=="0"]<-as.numeric(10)
df.f$CROPDMGEXP[(df.f$CROPDMGEXP)=="k"]<-as.numeric(10^3)
```

Here we create raw values and convert date field to date value.
```{r,cache=TRUE,tidy=FALSE}
library(lubridate)
library(dplyr)
df.m<-mutate(df.f,PROPDMG.2=PROPDMG*as.numeric(PROPDMGEXP),
             CROPDMG.2=CROPDMG*as.numeric(CROPDMGEXP),DATE=mdy_hms(BGN_DATE))
```

Here we remove one outlier field after exploring the data and only use data after 1995 as this is is regarded as more reliable data.
```{r,cache=TRUE,tidy=FALSE}
library(dplyr)
df.g<-mutate(df.m,TOTALDMG=PROPDMG.2+CROPDMG.2,YEAR = as.POSIXlt(DATE)$year + 1900) %>%
  filter(TOTALDMG<115032500000,YEAR>1995) %>%
  arrange(desc(TOTALDMG)) 
```
We now consolidata all the event to match NOAA documentation and to analyze data more easily. we use the case_when r function.  We also append the new values to the original data frame and create a new data frame newdata.

```{r,cache=TRUE,tidy=FALSE,echo=FALSE}
library(dplyr)
newevents<-case_when(
grepl("TORNADO",df.g$EVTYPE) ~ "Tornado",
grepl("FLOOD",df.g$EVTYPE) ~ "Flood",
grepl("HAIL",df.g$EVTYPE) ~ "Hail",
grepl("LIGHTNING",df.g$EVTYPE) ~ "Lightning",
grepl("HIGH WIND",df.g$EVTYPE)~ "High Wind" ,
grepl("STRONG WIND",df.g$EVTYPE)~"High Wind",
grepl("WINTER STORM",df.g$EVTYPE)~"Winter Storm",
grepl("HEAVY RAIN",df.g$EVTYPE)~"Heavy Rain" ,
grepl("HEAVY SNOW",df.g$EVTYPE)~"Heavy Snow",
grepl("WILDFIRE",df.g$EVTYPE)~"Wildfire",
grepl("URBAN/SML STREAM FLD",df.g$EVTYPE)~"Flood",
grepl("ICE STORM",df.g$EVTYPE)~"Ice Storm" ,
grepl("TSTM WIND/HAIL",df.g$EVTYPE)~"Hail",
grepl("TROPICAL STORM",df.g$EVTYPE)~"Tropical Storm",
grepl("WINTER WEATHER",df.g$EVTYPE)~"Winter Weather",
grepl("WILD/FOREST FIRE",df.g$EVTYPE)~"Wildfire",
grepl("DROUGHT",df.g$EVTYPE)~"Drought",
grepl("LAKE-EFFECT SNOW",df.g$EVTYPE)~"Lake-Effect Snow",
grepl("BLIZZARD",df.g$EVTYPE)~"Blizzard",
grepl("LANDSLIDE",df.g$EVTYPE)~"Heavy Rain",
grepl("STORM SURGE",df.g$EVTYPE)~"Storm Surge/Tide",
grepl("COASTAL FLOOD",df.g$EVTYPE)~"Coastal Flood",
grepl("WINTER WEATHER/MIX",df.g$EVTYPE)~"Winter Weather",
grepl("HURRICANE",df.g$EVTYPE)~"Hurricane (Typhoon)",
grepl("LIGHT SNOW",df.g$EVTYPE)~"Heavy Snow",
grepl("FROST/FREEZE",df.g$EVTYPE)~"Frost/Freeze",
grepl("MARINE TSTM WIND",df.g$EVTYPE)~"Marine Thunderstorm Wind",
grepl("RIVER FLOOD",df.g$EVTYPE)~"Flood",
grepl("EXTREME COLD",df.g$EVTYPE)~"Cold/Wind Chill",
grepl("DUST STORM",df.g$EVTYPE)~"Dust Storm",
grepl("DUST DEVIL",df.g$EVTYPE)~"Dust Devil",
grepl("HURRICANE/TYPHOON",df.g$EVTYPE)~"Hurricane (Typhoon)",
grepl("DRY MICROBURST",df.g$EVTYPE)~"Thunderstorm Wind",
grepl("FOG",df.g$EVTYPE)~"Dense Fog",
grepl("DENSE FOG",df.g$EVTYPE)~"Dense Fog",
grepl("AVALANCHE",df.g$EVTYPE)~"Avalanche",
grepl("HIGH SURF",df.g$EVTYPE)~"High Surf",
grepl("STORM SURGE/TIDE",df.g$EVTYPE)~"Storm Surge/Tide",
grepl("WIND",df.g$EVTYPE)~"Strong Wind",
grepl("TSTM WIND",df.g$EVTYPE)~"Thunderstorm Wind",
grepl("TROPICAL DEPRESSION",df.g$EVTYPE)~"Tropical Depression",
grepl("MARINE STRONG WIND",df.g$EVTYPE)~"Marine Strong Wind",
grepl("OTHER",df.g$EVTYPE)~"OTHER",
grepl("STRONG WINDS",df.g$EVTYPE)~"Strong Wind",
grepl("COASTAL FLOODING",df.g$EVTYPE)~"Coastal Flood",
grepl("WATERSPOUT",df.g$EVTYPE)~"Waterspout",
grepl("EXCESSIVE SNOW",df.g$EVTYPE)~"Heavy Snow",
grepl("EXCESSIVE HEAT",df.g$EVTYPE)~"Excessive Heat",
grepl("MARINE THUNDERSTORM WIND",df.g$EVTYPE)~"Marine Thunderstorm Wind",
grepl("LIGHT FREEZING RAIN",df.g$EVTYPE)~"Ice Storm",
grepl("Light Snow",df.g$EVTYPE)~"Heavy Snow",
grepl("EXTREME COLD/WIND CHILL",df.g$EVTYPE)~"xtreme Cold/Wind Chill",
grepl("HEAVY SURF",df.g$EVTYPE)~"High Surf",
grepl("ICY ROADS",df.g$EVTYPE)~"Ice Storm",
grepl("MARINE HIGH WIND",df.g$EVTYPE)~"Marine High Wind",
grepl("GUSTY WINDS",df.g$EVTYPE)~"Strong Wind",
grepl("HEAVY SURF/HIGH SURF",df.g$EVTYPE)~"High Surf",
grepl("COLD/WIND CHILL",df.g$EVTYPE)~"xtreme Cold/Wind Chill",
grepl("MIXED PRECIPITATION",df.g$EVTYPE)~"Heavy Rain",
grepl("TSUNAMI",df.g$EVTYPE)~"Tsunami",
grepl("GUSTY WIND",df.g$EVTYPE)~"Strong Wind",
grepl("FREEZE",df.g$EVTYPE)~"xtreme Cold/Wind Chill",
grepl("SNOW",df.g$EVTYPE)~"Heavy Snow",
grepl("Gusty Winds",df.g$EVTYPE)~"Strong Wind",
grepl("SMALL HAIL",df.g$EVTYPE)~"Hail",
grepl("SEICHE",df.g$EVTYPE)~"Seiche",
grepl("TSTM WIND",df.g$EVTYPE)~"Thunderstorm Wind",
grepl("TYPHOON",df.g$EVTYPE)~"Hurricane (Typhoon)",
grepl("ASTRONOMICAL HIGH TIDE",df.g$EVTYPE)~"Astronomical Low Tide",
grepl("FUNNEL CLOUD",df.g$EVTYPE)~"Funnel Cloud",
grepl("FREEZING FOG",df.g$EVTYPE)~"Freezing Fog",
grepl("FREEZING RAIN",df.g$EVTYPE)~"Frost/Freeze",
grepl("Coastal Flooding",df.g$EVTYPE)~"Coastal Flood",
grepl("Dust Devil",df.g$EVTYPE)~"Dust Devil",
grepl("HEAT",df.g$EVTYPE)~"Heat",
grepl("LAKESHORE FLOOD",df.g$EVTYPE)~"Lakeshore Flood",
grepl("RIP CURRENTS",df.g$EVTYPE)~"Rip Current",
grepl("River Flooding",df.g$EVTYPE)~"Flash Flood",
grepl("Coastal Flood",df.g$EVTYPE)~"Coastal Flood",
grepl("Cold",df.g$EVTYPE)~"Cold/Wind Chill",
grepl("EXTREME WINDCHILL",df.g$EVTYPE)~"Cold/Wind Chill",
grepl("AGRICULTURAL FREEZE",df.g$EVTYPE)~"Cold/Wind Chill",
grepl("COASTAL FLOODING/EROSION",df.g$EVTYPE)~"Coastal Flood",
grepl("Gradient wind",df.g$EVTYPE)~"Strong Wind",
grepl("LAKE EFFECT SNOW",df.g$EVTYPE)~"Lake-Effect Snow",
grepl("Mixed Precipitation",df.g$EVTYPE)~"Hail",
grepl("MUDSLIDE",df.g$EVTYPE)~"Debris Flow",
grepl("RAIN",df.g$EVTYPE)~"Heavy Rain",
grepl("SNOW SQUALLS",df.g$EVTYPE)~"Heavy Snow",
grepl("UNSEASONABLY COLD",df.g$EVTYPE)~"Cold/Wind Chill",
grepl("WET MICROBURST",df.g$EVTYPE)~"Thunderstorm Wind",
grepl("TSTM WIND",df.g$EVTYPE)~"Thunderstorm Wind",
grepl("ASTRONOMICAL LOW TIDE",df.g$EVTYPE)~"Astronomical Low Tide",
grepl("DAM BREAK",df.g$EVTYPE)~"Debris Flow",
grepl("Damaging Freeze",df.g$EVTYPE)~"Cold/Wind Chill",
grepl("Erosion/Cstl Flood",df.g$EVTYPE)~"Flash Flood",
grepl("Freeze",df.g$EVTYPE)~"xtreme Cold/Wind Chill",
grepl("Freezing Drizzle",df.g$EVTYPE)~"xtreme Cold/Wind Chill",
grepl("FREEZING DRIZZLE",df.g$EVTYPE)~"xtreme Cold/Wind Chill",
grepl("Glaze",df.g$EVTYPE)~"OTHER",
grepl("gradient wind",df.g$EVTYPE)~"High Wind",
grepl("High Surf",df.g$EVTYPE)~"High Surf",
grepl("HIGH WIND",df.g$EVTYPE)~"High Wind",
grepl("LANDSPOUT",df.g$EVTYPE)~"Debris Flow",
grepl("MARINE HAIL",df.g$EVTYPE)~"Marine Hail",
grepl("MUD SLIDE",df.g$EVTYPE)~"Debris Flow",
grepl("Snow",df.g$EVTYPE)~"Heavy Snow",
grepl("Snow Squalls",df.g$EVTYPE)~"Heavy Snow",
grepl("Strong Wind",df.g$EVTYPE)~"Strong Wind",
grepl("TIDAL FLOODING",df.g$EVTYPE)~"Storm Surge/Tide",
grepl("UNSEASONAL RAIN",df.g$EVTYPE)~"Heavy Rain",
grepl("VOLCANIC ASH",df.g$EVTYPE)~"Volcanic Ash",
grepl("	   HIGH SURF ADVISORY",df.g$EVTYPE)~"OTHER",
grepl("FLASH FLOOD",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND (G45)",df.g$EVTYPE)~"OTHER",
grepl("Beach Erosion",df.g$EVTYPE)~"OTHER",
grepl("BLOWING DUST",df.g$EVTYPE)~"OTHER",
grepl("blowing snow",df.g$EVTYPE)~"OTHER",
grepl("COASTAL  FLOODING/EROSION",df.g$EVTYPE)~"OTHER",
grepl("COASTAL EROSION",df.g$EVTYPE)~"OTHER",
grepl("Coastal Storm",df.g$EVTYPE)~"OTHER",
grepl("COLD",df.g$EVTYPE)~"OTHER",
grepl("DAMAGING FREEZE",df.g$EVTYPE)~"OTHER",
grepl("DENSE SMOKE",df.g$EVTYPE)~"OTHER",
grepl("DOWNBURST",df.g$EVTYPE)~"OTHER",
grepl("Early Frost",df.g$EVTYPE)~"OTHER",
grepl("Extended Cold",df.g$EVTYPE)~"OTHER",
grepl("Extreme Cold",df.g$EVTYPE)~"OTHER",
grepl("FLASH FLOOD/FLOOD",df.g$EVTYPE)~"OTHER",
grepl("FLOOD/FLASH/FLOOD",df.g$EVTYPE)~"OTHER",
grepl("Freezing drizzle",df.g$EVTYPE)~"OTHER",
grepl("Freezing Rain",df.g$EVTYPE)~"OTHER",
grepl("FROST",df.g$EVTYPE)~"OTHER",
grepl("Frost/Freeze",df.g$EVTYPE)~"OTHER",
grepl("GLAZE",df.g$EVTYPE)~"OTHER",
grepl("GRADIENT WIND",df.g$EVTYPE)~"OTHER",
grepl("GUSTY WIND/HAIL",df.g$EVTYPE)~"OTHER",
grepl("GUSTY WIND/HVY RAIN",df.g$EVTYPE)~"OTHER",
grepl("Gusty wind/rain",df.g$EVTYPE)~"OTHER",
grepl("HARD FREEZE",df.g$EVTYPE)~"OTHER",
grepl("Heavy Rain/High Surf",df.g$EVTYPE)~"OTHER",
grepl("Heavy snow shower",df.g$EVTYPE)~"OTHER",
grepl("Heavy Surf",df.g$EVTYPE)~"OTHER",
grepl("HIGH SEAS",df.g$EVTYPE)~"OTHER",
grepl("HIGH SWELLS",df.g$EVTYPE)~"OTHER",
grepl("HIGH WINDS",df.g$EVTYPE)~"OTHER",
grepl("Ice jam flood \\(minor",df.g$EVTYPE)~"OTHER",
grepl("ICE ROADS",df.g$EVTYPE)~"OTHER",
grepl("Lake Effect Snow",df.g$EVTYPE)~"OTHER",
grepl("LANDSLIDES",df.g$EVTYPE)~"OTHER",
grepl("Landslump",df.g$EVTYPE)~"OTHER",
grepl("LATE SEASON SNOW",df.g$EVTYPE)~"OTHER",
grepl("Light snow",df.g$EVTYPE)~"OTHER",
grepl("Light Snowfall",df.g$EVTYPE)~"OTHER",
grepl("Marine Accident",df.g$EVTYPE)~"OTHER",
grepl("Microburst",df.g$EVTYPE)~"OTHER",
grepl("NON-SEVERE WIND DAMAGE",df.g$EVTYPE)~"OTHER",
grepl("NON-TSTM WIND",df.g$EVTYPE)~"OTHER",
grepl("Other",df.g$EVTYPE)~"OTHER",
grepl("RIP CURRENT",df.g$EVTYPE)~"OTHER",
grepl("ROCK SLIDE",df.g$EVTYPE)~"OTHER",
grepl("ROUGH SURF",df.g$EVTYPE)~"OTHER",
grepl("SNOW SQUALL",df.g$EVTYPE)~"OTHER",
grepl("Strong Winds",df.g$EVTYPE)~"OTHER",
grepl("Tidal Flooding",df.g$EVTYPE)~"OTHER",
grepl("Tstm Wind",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND  (G45)",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND (41)",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND (G35)",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND 40",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND 45",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND AND LIGHTNING",df.g$EVTYPE)~"OTHER",
grepl("TSTM WIND G45",df.g$EVTYPE)~"OTHER",
grepl("Unseasonable Cold",df.g$EVTYPE)~"OTHER",
grepl("UNSEASONABLY WARM",df.g$EVTYPE)~"OTHER",
grepl("Whirlwind",df.g$EVTYPE)~"OTHER",
grepl("WHIRLWIND",df.g$EVTYPE)~"OTHER",
grepl("Wind",df.g$EVTYPE)~"OTHER",
grepl("WIND AND WAVE",df.g$EVTYPE)~"OTHER",
grepl("Wind Damage",df.g$EVTYPE)~"OTHER",
grepl("WINTER WEATHER MIX",df.g$EVTYPE)~"OTHER",
grepl("Wintry Mix",df.g$EVTYPE)~"OTHER",
grepl("WINTRY MIX",df.g$EVTYPE)~"OTHER"
)

newevents<-as.data.frame(newevents)

newdata<-cbind(df.g,newevents)

newdata<-mutate(newdata,newevents=as.character(newevents))

```
```{r}
dim(newdata)
names(newdata)
head(newdata)
```

# Results

## Most Harmful Events - Population Health

In order to determine which events accross the United States are most harmful with respect to population health we combine both fatalities and injuries.
Based on the 

```{r,cache=TRUE,tidy=FALSE}

library(dplyr)
library(ggplot2)

df.harm<-mutate(newdata,TotalHarm=FATALITIES+INJURIES)
df.harm2<-group_by(df.harm,newevents) %>%
  summarize(TotalHarmSum=sum(TotalHarm)) %>%
  arrange(desc(TotalHarmSum))

head(df.harm2,20)

df.harm2.f<-filter(df.harm2,TotalHarmSum>507)

ggplot(df.harm2.f,aes(newevents,TotalHarmSum))+geom_bar(stat = "identity")+ 
  theme(axis.text.x = element_text(angle = 45,hjust = 1))

```

From the chart above we see that Tornado is the most impact on population health in the United States.

We also see based on yearly data that one single major event can contribute to a high number of injuries and fatalities.

```{r,cache=TRUE,tidy=FALSE}
library(dplyr)
par(mfrow=c(2,2))

tornado<-filter(df.harm,newevents=="Tornado") %>%
  group_by(YEAR) %>%
  summarize(TornadoHarm=sum(TotalHarm))
with(tornado,plot(YEAR,TornadoHarm,type="l"))

flood<-filter(df.harm,newevents=="Flood") %>%
  group_by(YEAR) %>%
  summarize(FloodHarm=sum(TotalHarm))
with(flood,plot(YEAR,FloodHarm,type="l"))

StrngWnd<-filter(df.harm,newevents=="Strong Wind") %>%
  group_by(YEAR) %>%
  summarize(StrngWndHarm=sum(TotalHarm))
with(StrngWnd,plot(YEAR,StrngWndHarm,type="l"))

Canes<-filter(df.harm,newevents=="Hurricane (Typhoon)") %>%
  group_by(YEAR) %>%
  summarize(HurricaneHarm=sum(TotalHarm))
with(Canes,plot(YEAR,HurricaneHarm,type="l"))

```

## Most Economical Impactful Events

Across the United Sates Hurricane, Flood, Storm Surges, and Tornados have the greatest economic impact being heavily weighted by property damage cost with Hurricane causing 87 Billion dollars in Damage since 1996.




```{r,cache=TRUE,tidy=FALSE}
library(dplyr)
library(ggplot2)

df.g3<-group_by(newdata,newevents) %>%
  summarize(TOTAL_CROPDMG=sum(CROPDMG.2),TOTAL_PROPDMG=sum(PROPDMG.2),TOTAL=sum(TOTALDMG)) %>%
  arrange(desc(TOTAL))
head(df.g3,20)

df.g3.f20<-filter(df.g3,TOTAL >= 207835000)

ggplot(df.g3.f20,aes(newevents,TOTAL))+geom_bar(stat = "identity")+ 
  theme(axis.text.x = element_text(angle = 45,hjust = 1))
```

We also see based on yearly data that one single major event can contribute to majority of ecomonomical impact acrued throughtout the years

```{r,cache=TRUE,tidy=FALSE}
library(dplyr)

par(mfrow=c(2,2))

canes<-filter(newdata,newevents=="Hurricane (Typhoon)") %>%
  group_by(YEAR) %>%
  summarize(HurricaneDamage=sum(TOTALDMG))
with(canes,plot(YEAR,HurricaneDamage,type="l"))

flood<-filter(newdata,newevents=="Flood") %>%
  group_by(YEAR) %>%
  summarize(FloodDamage=sum(TOTALDMG))
with(flood,plot(YEAR,FloodDamage,type="l"))

surge<-filter(newdata,newevents=="Storm Surge/Tide") %>%
  group_by(YEAR) %>%
  summarize(StormSurgeDamage=sum(TOTALDMG))
with(surge,plot(YEAR,StormSurgeDamage,type="l"))

tornado<-filter(newdata,newevents=="Tornado") %>%
  group_by(YEAR) %>%
  summarize(TornadoDamage=sum(TOTALDMG))
with(tornado,plot(YEAR,TornadoDamage,type="l"))

```

# Conclusion

Based on this analyis Tornados are the most harmful events with respect to public health suggesting these events are more unpredictable with short warning  systems.  Areas prones to tornadoes should have more safe houses and strong building codes.   As far as economical impact coastal areas in  eastern seaboard  and gulf coast particularly in Florida are the most impactful due to hurricanes suggesting, in additional recommendation above,  possible addtional insurance funds should be raised to save for a rainy day.
 
 














